---
output: github_document
---

mrgPBPK is an R package that provides a set of functions to be used for PBPK modeling. The functions mainly generate drug- and system-specific parameters required to build a PBPK model. There are two versions of the system-specific parameter-generating functions: a general PBPK model version and a monoclonal Ab PBPK model version.

# Load libraries

```{r setup, warning=FALSE, message=FALSE}
library(mrgPBPK)
library(dplyr)
library(ggplot2)
library(cowplot)
library(GGally)
```

# Generating drug-specific parameters

## Calculate tissue:plasma partition coefficients (Kp)

The function `calcKp` can calculate the molecule's Kp values for different organs using one of five different calculation methods:

- PT: Poulin and Theil 

```{r}
# define molecule's physicochemical properties
logP <- 2  #lipophilicity
pKa <- 1  #acidic strength
type <- 3  #type of molecule
BP <- 1  #blood:plasma concentration ratio
fup <- 0.5  #unbound fraction in plasma
method <- "PT"  #prediction method

# calculate partition coefficients
Kps <- calcKp(logP=logP, pKa=pKa, fup=fup, BP=BP, type=type, method=method)
df_Kps <- bind_rows(Kps)
df_Kps2 <- tibble(Parameter=names(df_Kps), Value=as.numeric(Kps))

ggplot(data=df_Kps2, aes(Parameter, Value)) +
  geom_col() + theme_bw()
```
## Calculate blood:plasma concentration ratio (BP)

In case BP parameter was missing, the function `calcBP` can be used 

```{r}
# in case BP parameter was missing, the function calcBP
```


## Calculate unbound fraction in plasma (fup)

# General PBPK modeling

## Generate individual physiological parameters

```{r}
# define the individual demographics
## You only need to define two of the three covariates: body weight, height, and BMI
age <- 30
ismale <- TRUE
bw <- 73
ht <- 1.76

# generate individual physiological parameters
indPars <- genInd(age=age, is.male=ismale, bw_targ=bw, ht_targ=ht, optimize = FALSE)
df_indPars <- bind_rows(indPars) %>% select(contains(c("Q","V")))
df_indPars2 <- tibble(Parameter=names(df_indPars), Value=as.numeric(df_indPars))

plot_vols <- ggplot(data=df_indPars2 %>% filter(grepl("V", Parameter)), aes(Parameter, Value)) + geom_col() + theme_bw() + labs(title = "Volumes")
plot_flows <- ggplot(data=df_indPars2 %>% filter(grepl("Q", Parameter)), aes(Parameter, Value)) + geom_col() + theme_bw() + labs(title = "Flows")

plot_grid(plot_vols, plot_flows, ncol=1)
```

## Generate population physiological parameters

```{r}
# define the population demographics
## You only need to define the range for two of the three covariates: body weight, height, and BMI
nSubj <- 40  #number of subjects
minAge <- 20  #minimum age
maxAge <- 80  #maximum age
femPerc <- 50  #percentage of females
minBW <- 50  #minimum body weight
maxBW <- 100  #maximum body weight
minHT <- 1.5  #minimum height
maxHT <- 1.9  #maximum height

# generate population physiological parameters
popPars <- genPop(nSubj=nSubj, minAge=minAge, maxAge=maxAge, femPerc=femPerc, minBW=minBW, maxBW=maxBW, minHT=minHT, maxHT=maxHT, optimize = FALSE)
df_popPars <- bind_rows(popPars) %>% select(ID, everything())

df_popPars <- bind_rows(popPars) %>%
  mutate(SEX = ifelse(SEX == 1, "male", "female"),
         SEX = factor(SEX))
df_vols <- df_popPars %>% 
  select(AGE, HT, BW, SEX, contains("V"))

df_flows <- df_popPars %>% 
  select(AGE, HT, BW, SEX, contains("Q"))

plot_corr_vols <- ggcorr(df_vols %>% select(-SEX, -AGE, -BW, -HT, -Vot)) + labs(title = "Volumes")
plot_corr_flows <- ggcorr(df_flows %>% select(-SEX, -AGE, -BW, -HT, -Qot)) + labs(title = "Flows")
plot_grid(plot_corr_vols, plot_corr_flows, ncol=2)
```

# Monoclonal antibody modeling

## Generate individual physiological parameters

```{r}
# define the individual demographics
## You only need to define two of the three covariates: body weight, height, and BMI
age <- 30
ismale <- T
bw <- 73
ht <- 1.76

# generate individual physiological parameters
indPars <- genInd_mab(age=age, is.male=ismale, bw_targ=bw, ht_targ=ht)
df_indPars <- bind_rows(indPars)

head(t(df_indPars), 10)
```


## Generate population physiological parameters

```{r}
# define the population demographics
## You only need to define the range for two of the three covariates: body weight, height, and BMI
nSubj <- 40  #number of subjects
minAge <- 20  #minimum age
maxAge <- 80  #maximum age
femPerc <- 50  #percentage of females
minBW <- 50  #minimum body weight
maxBW <- 100  #maximum body weight
minHT <- 1.5  #minimum height
maxHT <- 1.9  #maximum height

# generate population physiological parameters
popPars <- genPop_mab(nSubj=nSubj, minAge=minAge, maxAge=maxAge, femPerc=femPerc, minBW=minBW, maxBW=maxBW, minHT=minHT, maxHT=maxHT)
df_popPars <- bind_rows(popPars) %>% select(ID, everything())

summary(df_popPars[,1:11])
```


