---
output: github_document
---

# mrgPBPK

mrgPBPK is an open-source R package that provides a set of functions to be used for PBPK modeling. The functions mainly generate drug- and system-specific parameters required to build a PBPK model. 

# Load libraries

```{r setup, warning=FALSE, message=FALSE}
library(mrgPBPK)
library(dplyr)
library(ggplot2)
library(cowplot)
library(GGally)
```

# Generate drug-specific parameters

## Calculate tissue:plasma partition coefficients (Kp)

The function `calcKp` can calculate the molecule's Kp values for different organs using one of five different calculation methods:

- PT: Poulin and Theil ([source](https://pubmed.ncbi.nlm.nih.gov/11782904/)).
- RR: Rodgers and Rowland ([source1](https://pubmed.ncbi.nlm.nih.gov/15858854/) and [source2](https://pubmed.ncbi.nlm.nih.gov/16639716/)).
- Berez: Berezhkovskiy ([source](https://pubmed.ncbi.nlm.nih.gov/15124219/)).
- Schmitt: Schimtt ([source](https://pubmed.ncbi.nlm.nih.gov/17981004/)).
- pksim: PK-Sim ([source](https://www.tandfonline.com/doi/abs/10.1517/17425255.1.1.159)).

The function uses the unified tissue composition data reported in https://dmd.aspetjournals.org/content/48/10/903

```{r}
# define molecule's physicochemical properties
logP <- 2  #lipophilicity
pKa <- 1  #acidic strength
type <- 3  #type of molecule
BP <- 1  #blood:plasma concentration ratio
fup <- 0.5  #unbound fraction in plasma
method <- "PT"  #prediction method

# calculate partition coefficients
Kps <- calcKp(logP=logP, pKa=pKa, fup=fup, BP=BP, type=type, method=method)
df_Kps <- bind_rows(Kps)
df_Kps2 <- tibble(Parameter=names(df_Kps), Value=as.numeric(Kps))

ggplot(data=df_Kps2, aes(Parameter, Value)) +
  geom_col() + theme_bw()
```

## Calculate blood:plasma concentration ratio (BP)

In case BP parameter was missing, the function `calcBP` can be used to calculate BP based on the methods reported [here](https://pubmed.ncbi.nlm.nih.gov/20549836/). There are two methods to chose from:

- Method 1: uses molecule's fup to calculate BP.
- Method 2: uses molecule's logP (or another measurement of lipophilicity like logD) to calculate BP.

Note: drug type = "total" is the default type and it uses the regression coefficients calculated by fitting different molecule types (acids, bases, and neutrals) together.

```{r}
# in case BP parameter was missing, the function calcBP
calcBP(fup = fup, method = 1)
calcBP(logP = logP, fup = fup, method = 2)
```

## Calculate unbound fraction in plasma (fup)

In case fup parameter was missing, the function `calcFup` can be used to calculate fup using the molecule's logP (or another measurement of lipophilicity like logD) based on the method reported [here](https://pubmed.ncbi.nlm.nih.gov/20549836/). 

```{r}
# in case BP parameter was missing, the function calcBP
calcFup(logP = logP)
```

# Generate system-specific parameters

System-related (physiologic) parameters required for PBPK modeling are the organ volumes and blood flows. There are two versions of the functions that generate the physiologic parameters, a version to be applied for general PBPK modeling of small molecules and another specific for monoclonal antibody PBPK modeling.  

## Small molecule PBPK modeling

Two functions `genInd` and `genPop` can be used to generate a virtual individual or a population, respectively, for PBPK modeling of small molecules. The algorithm used is adapted from Willmann et al 2007 [source](https://pubmed.ncbi.nlm.nih.gov/17431751/) and it maintains a correlation between the physiologic parameters and the sampled individual's covariates to generate realistic individuals. The algorithm is guided by these databases:

- NHANES data for anthropometric data ([source](https://www.cdc.gov/nchs/nhanes/index.htm)).
- ICRP data for physiologic parameters of typical individuals ([source](https://journals.sagepub.com/doi/pdf/10.1177/ANIB_32_3-4)).
- Willmann et al 2007 data for organ variability ([source](https://pubmed.ncbi.nlm.nih.gov/17431751/)).

The function `genInd` takes age, sex, and two inputs of body weight, height, and BMI. It returns a named list of the covariates and the physiologic parameters with volumes prefixed with `V` and flows prefixed with `Q`. The logical argument `optimize` states if the user wants to run an additional optimization step on the generated organ volumes. This step would give better parameter predictions but will take longer to run. 

```{r}
# define the individual demographics
## You only need to define two of the three covariates: body weight, height, and BMI
age <- 30
ismale <- TRUE
bw <- 73
ht <- 1.76

# generate individual physiological parameters
indPars <- genInd(age=age, is.male=ismale, bw_targ=bw, ht_targ=ht, optimize = FALSE)
df_indPars <- bind_rows(indPars) %>% select(contains(c("Q","V")))
df_indPars2 <- tibble(Parameter=names(df_indPars), Value=as.numeric(df_indPars))

plot_vols <- ggplot(data=df_indPars2 %>% filter(grepl("V", Parameter)), aes(Parameter, Value)) + geom_col() + theme_bw() + labs(title = "Volumes")
plot_flows <- ggplot(data=df_indPars2 %>% filter(grepl("Q", Parameter)), aes(Parameter, Value)) + geom_col() + theme_bw() + labs(title = "Flows")

plot_grid(plot_vols, plot_flows, ncol=1)
```

The function `genPop` takes ranges of ages, and two ranges of body weights, heights, or BMIs as well as the percentage of females in the desired population. It returns a named list of lists with the upper level being the individuals' IDs and the nested list is each individual's covariates and physiologic parameters.

```{r}
# define the population demographics
## You only need to define the range for two of the three covariates: body weight, height, and BMI
nSubj <- 40  #number of subjects
minAge <- 20  #minimum age
maxAge <- 80  #maximum age
femPerc <- 50  #percentage of females
minBW <- 50  #minimum body weight
maxBW <- 100  #maximum body weight
minHT <- 1.5  #minimum height
maxHT <- 1.9  #maximum height

# generate population physiological parameters
popPars <- genPop(nSubj=nSubj, minAge=minAge, maxAge=maxAge, femPerc=femPerc, minBW=minBW, maxBW=maxBW, minHT=minHT, maxHT=maxHT, optimize = FALSE)
df_popPars <- bind_rows(popPars) %>% select(ID, everything())

df_popPars <- bind_rows(popPars) %>%
  mutate(SEX = ifelse(SEX == 1, "male", "female"),
         SEX = factor(SEX))
df_vols <- df_popPars %>% 
  select(AGE, HT, BW, SEX, contains("V"))

df_flows <- df_popPars %>% 
  select(AGE, HT, BW, SEX, contains("Q"))

plot_corr_vols <- ggcorr(df_vols %>% select(-SEX, -AGE, -BW, -HT, -Vot)) + labs(title = "Volumes")
plot_corr_flows <- ggcorr(df_flows %>% select(-SEX, -AGE, -BW, -HT, -Qot)) + labs(title = "Flows")
plot_grid(plot_corr_vols, plot_corr_flows, ncol=2)
```

# Monoclonal antibody PBPK modeling

The functions `genInd_mab` and `genPop_mab` take the same inputs as `genInd` and `genPop`, respectively, but they return physiologic parameters to be used for monoclonal antibody (mAb) PBPK modeling. The parameter names are consistent with the names used in the mAb PBPK model developed [here](https://github.com/metrumresearchgroup/bioPBPK/tree/main/mAb_bamlanivimab) and was based on the model reported by Shah and Betts 2012 ([source](https://pubmed.ncbi.nlm.nih.gov/22143261/)) and Jones et al 2019 ([source](https://ascpt.onlinelibrary.wiley.com/doi/full/10.1002/psp4.12461)).  

```{r}
# define the individual demographics
## You only need to define two of the three covariates: body weight, height, and BMI
age <- 30
ismale <- T
bw <- 73
ht <- 1.76

# generate individual physiological parameters
indPars <- genInd_mab(age=age, is.male=ismale, bw_targ=bw, ht_targ=ht)
df_indPars <- bind_rows(indPars)

head(t(df_indPars), 10)
```

## Generate population physiological parameters

```{r}
# define the population demographics
## You only need to define the range for two of the three covariates: body weight, height, and BMI
nSubj <- 40  #number of subjects
minAge <- 20  #minimum age
maxAge <- 80  #maximum age
femPerc <- 50  #percentage of females
minBW <- 50  #minimum body weight
maxBW <- 100  #maximum body weight
minHT <- 1.5  #minimum height
maxHT <- 1.9  #maximum height

# generate population physiological parameters
popPars <- genPop_mab(nSubj=nSubj, minAge=minAge, maxAge=maxAge, femPerc=femPerc, minBW=minBW, maxBW=maxBW, minHT=minHT, maxHT=maxHT)
df_popPars <- bind_rows(popPars) %>% select(ID, everything())

summary(df_popPars[,1:11])
```


